---
title: "GIS 3 Data processing for original input files"
output: html_notebook
---

# Purpose of file
This is the very first file in the coding workflow for the project. I've gathered data on pokemon spawn points, census tract polygons, census tract attributes, locations of restaurants, etc.

But these need extensive data cleaning and some spatial computations (e.g. to get total number of bike trip destinations by location). The results will be saved in GIS3SpatialRawData folder.

# Step 0: Libraries
```{r}
library(tidyverse)
library(sf)
library(RColorBrewer)
library(tmap)
setwd("~/Desktop/GIS3FinalProject")
```

# Step 1: Process pokemon points
## Step 1.1 Load in raw pokemon data
```{r}
pokemon_raw <- read.csv("./GIS3RawData/pokemon-spawns.csv")
head(pokemon_raw)
str(pokemon_raw)
```
## Step 1.2: Process raw pokemon data to have spatial long lat, and count number of pokemons at each point
```{r}
pokemon_processing <- pokemon_raw
colnames(pokemon_processing)

# Convert string long lats to numeric long lats
pokemon_processing$lat <- as.numeric(pokemon_processing$lat)
pokemon_processing$lng <- as.numeric(pokemon_processing$lng)
pokemon_processing

# Count number of pokemons spawned at each long lat point of pokemon spawn
pokemon_point_counts <-
  pokemon_processing %>% 
  group_by(lat, lng) %>%
  summarize(NumPokemon = n()) %>%
  rename(long='lng') %>%
  arrange(long, lat)

# Create ID for each long lat
pokemon_point_counts$AllPokeLocID <- 1:nrow(pokemon_point_counts)

# Relocate columns
pokemon_point_counts <-
  pokemon_point_counts %>%
  relocate(AllPokeLocID, .before=lat) %>%
  relocate(lat, .after=long)

# Sanity check
pokemon_point_counts %>% arrange(long)
pokemon_point_counts %>% arrange(-long)

# Convert long lat csv to shapefile 
pokemon_point_counts_sf <- st_as_sf(pokemon_point_counts, coords = c("long", "lat"), crs=4326) 
```

# Step 2: Process San Francisco 2010 census tracts
## 2.1 Load polygon file of census tracts
```{r}
SanFran_tracts <- st_read('./GIS3RawData/SF2010Tracts.geojson')

head(SanFran_tracts)
```
## 2.2 Setting EPSGs and joining pokemon spawn points with counts to San Francisco tracts so that each row (point) is assigned a tract
```{r}
# Convert both census tract points and census tracts to epsg 7131
pokemon_point_counts_7131 <- pokemon_point_counts_sf %>% st_transform(7131)
SanFran_tracts_7131 <- SanFran_tracts %>% st_transform(7131)

# Spatial join to give each point a tract
pokemon_point_counts_SanFran <-
  st_join(pokemon_point_counts_7131, 
          SanFran_tracts_7131, join = st_within, left=FALSE) %>%
  select(AllPokeLocID, NumPokemon, geometry, geoid10)

pokemon_point_counts_SanFran$geoid10 <- as.numeric(pokemon_point_counts_SanFran$geoid10)
pokemon_point_counts_SanFran <-
  pokemon_point_counts_SanFran %>% rename(GeoFIPS = 'geoid10')

head(pokemon_point_counts_SanFran) 
```
## 2.3 Find only relevant tracts that have pokemon points
```{r}
# Union individual points to one object so that we can intersect tracts with this point object
unioned_pokemon_point_counts_SanFran <-
  st_union(pokemon_point_counts_SanFran)

# Select only tracts that actually have points
relevant_SanFran_tracts_7131 <-
  SanFran_tracts_7131 %>%
  filter(st_intersects(., unioned_pokemon_point_counts_SanFran, sparse=FALSE)) %>%
  mutate(GeoFIPS = as.numeric(geoid10))

relevant_SanFran_tracts_7131
```

## 2.4 Quick tmap visualization of number of pokemon at each point, and only in census tracts that actually have points
```{r}
#SanFran_tracts_7132

nrow(relevant_SanFran_tracts_7131)

tm_shape(SanFran_tracts_7131) +
  tm_polygons(alpha=0.5) +
tm_shape(pokemon_point_counts_SanFran) +
  tm_dots(col='NumPokemon', pal='BuPu')

tm_shape(relevant_SanFran_tracts_7131) +
  tm_polygons(alpha=0.5) +
tm_shape(pokemon_point_counts_SanFran) +
  tm_dots(col='NumPokemon', pal='BuPu')
```
# Step 3: Wrangle Census data
## 3.1 Load in raw census socioeconomic data
```{r}
SanFran_ACS_raw <- read.csv("./GIS3RawData/SanFranACS.csv")
SanFran_ACS_raw
```

## 3.2 Rename and mutate census variables to more intuitive names
```{r}
SanFran_ACS_renamed <-
  SanFran_ACS_raw %>% 
  select(Geo_FIPS, Geo_GEOID, Geo_NAME,
         tpop = SE_A01001_001,
         pden = SE_A00002_002,
         apop = SE_A03001_005,
         bpop = SE_A03001_003,
         hpop = SE_A04001_010,
         uemp = SE_A17005_003 / SE_A17005_001,
         pinc = SE_A14024_001,
         ag_rent = SE_A18005_001, # avg gross rent
         commute = SE_A09003_001 # commute time
         ) %>%
  mutate(aprp = apop/tpop, bprp = bpop/tpop, hprp = hpop / tpop) %>%
  mutate(minoprp = aprp + bprp + hprp) %>%
  relocate(minoprp, .before= aprp) %>%
  select(-apop, -bpop, -hpop) %>%
  rename(GeoFIPS = 'Geo_FIPS') %>%
  mutate(GeoFIPS = as.numeric(GeoFIPS)) %>%
  filter(GeoFIPS %in% relevant_SanFran_tracts_7132$GeoFIPS)

SanFran_ACS_renamed
```
## 3.3 Merge pokemon points with census tract attributes with the census tract data, so that each point is assigned census tract variable columns (e.g. population density)
```{r}
pokemon_point_counts_SanFran_ACS <-
  merge(pokemon_point_counts_SanFran, SanFran_ACS_renamed, all.x=TRUE,
        by='GeoFIPS') 
  # a few census tracts without ACS

length(unique(pokemon_point_counts_SanFran$GeoFIPS))

data.frame(pokemon_point_counts_SanFran_ACS) %>% arrange(-GeoFIPS)

data.frame(pokemon_point_counts_SanFran_ACS) %>% drop_na()
```

# Step 4: Cleaning data on location of "posh" restaurants in San Fran
Posh restaurants include boba shops, coffee shops, etc. Not just usual cheap restaurants.

## 4.1 Load posh restaurant data
```{r}
posh_restaurants_raw <- read.csv("./GIS3RawData/core_poi.csv")

posh_restaurants_raw
```

## 4.2 Convert posh restaraunt data into spatial format
```{r}
posh_restaurants_processing <- posh_restaurants_raw
posh_restaurants_processing <- 
  posh_restaurants_processing %>%
  select(placekey, location_name, lat = latitude, long=longitude)

posh_restaurants_points <-
   st_as_sf(posh_restaurants_processing, coords=c('long','lat'), crs=4326)
```

## Step 5: Clean data on general restaurants in San Fran

## 5.1 Load general restaurant data
```{r}
restaurants_raw <-
  read.csv("./GIS3RawData/restaurant-scores-lives-standard.csv")
```

## 5.2 Convert restaurant csv data to points via long lat
```{r}
restaurants_processing <- 
  restaurants_raw %>% 
  select(business_id, business_name, long = business_longitude, lat = business_latitude) %>%
  drop_na()

restaurants_points <- 
  st_as_sf(restaurants_processing, coords=c('long','lat'), crs=4326)

head(data.frame(restaurants_points))
tm_shape(relevant_SanFran_tracts_7132) +
  tm_polygons(alpha=0.5) +
tm_shape(restaurants_points) +
  tm_dots(col='blue')
```

# Step 6: Process traffic density (San Fran Lyft wheel bike rides) data

## 6.1 Load in the traffic density data
```{r}
bike_trips <- read.csv("./GIS3RawData/201906-baywheels-tripdata.csv")
head(bike_trips)
```

## 6.2 Aggregate trips to trips by destination, and convert to csv
```{r}
trips_by_destination = 
  bike_trips %>% 
  group_by(end_station_id, end_station_longitude, end_station_latitude) %>% 
  summarize(num_trips = n()) %>%
  rename(end_long = 'end_station_longitude', end_lat = 'end_station_latitude')

trips_by_destination

trips_by_destination_points <- 
  st_as_sf(trips_by_destination, coords=c('end_long','end_lat'), crs=4326)

tm_shape(relevant_SanFran_tracts_7132) +
  tm_polygons(alpha=0.5) +
tm_shape(trips_by_destination_points) +
  tm_dots(col='red')
```

# Step 7: Wrangling college location data
## 7.1 Load college data
```{r}
colleges_raw <- read.csv("./GIS3RawData/Map_of_Colleges_in_San_Francisco__2011_.csv")
head(colleges_raw)
```

## 7.2 Convert college location data into spatial long lat format
```{r}
colleges_processing <- colleges_raw
str(colleges_processing)

# String wrangling to convert Location column into long lat
colleges_processing$Location <-
  gsub("\\(", "", colleges_processing$Location)
colleges_processing$Location <-
  gsub("\\)", "", colleges_processing$Location)
colleges_processing$Location <-
  gsub(" ", "", colleges_processing$Location)

colleges_processing

colleges_processing <-
  separate(data = colleges_processing,
           col = Location, into = c("lat", "long"), sep = ",")

colleges_processing$long <- as.numeric(as.character(colleges_processing$long))
colleges_processing$lat <- as.numeric(as.character(colleges_processing$lat))

colleges_processing <-
  colleges_processing %>% select(ID, Institution, long, lat) %>% drop_na()

head(colleges_processing)
summary(colleges_processing)

# Final save of college points
colleges_points <-
  st_as_sf(colleges_processing, coords = c('long', 'lat'), crs = 4326) %>% 
  st_transform(4326)

head(data.frame(colleges_points))

plot(colleges_points)
```

# Step 8: Transform all cleaned data files into EPSG 4326
```{r}
# Polygons of tracts that contain pokemon spawn points
relevant_SanFran_tracts_4326 <-
  relevant_SanFran_tracts_7131 %>% st_transform(4326)

st_write(relevant_SanFran_tracts_4326, "./GIS3SpatialRawData/RelevantSanFranTracts4326.geojson")

# Polygons of pokemon spawn points with pokemon spawn count and census tract attributes
pokemon_point_counts_SanFran_ACS <-
  pokemon_point_counts_SanFran_ACS %>% st_transform(4326)

st_write(pokemon_point_counts_SanFran_ACS, "./GIS3SpatialRawData/PokemonPointsACS.geojson")

# Point locations of "posh" restaurants like boba shops and cafes
posh_restaurants_points <-
  posh_restaurants_points %>% st_transform(4326)

st_write(posh_restaurants_points, "./GIS3SpatialRawData/PoshRestaurantPoints.geojson")

# Point locations of general restaurants
restaurants_points <-
  restaurants_points %>% st_transform(4326)

st_write(restaurants_points, "./GIS3SpatialRawData/RestaurantPoints.geojson")

# Point locations of Lyft wheel trip destinations, with count by destination of trip as an attribute
trips_by_destination_points <-
  trips_by_destination_points %>% st_transform(4326)

st_write(trips_by_destination_points,
         "./GIS3SpatialRawData/TripsByDestination.geojson")

# Locations of colleges in San Fran
colleges_points <-
  colleges_points %>% 
  st_transform(4326)

st_write(colleges_points, "./GIS3SpatialRawData/CollegePoints.geojson")
```
